Combining living and fossil taxa into phylogenies: the missing data issue

### 2014-05-21 Manuscript/TEM_draft.tex
Added RPBC test

For the Bayesian trees, we had to transform the trees posterior distribution to a single value in order to be able to compare it to the results of the Maximum Likelihood trees comparisons.
Traditionally people use the 50\% consensus rule in order to summarise a Bayesian tree posterior distribution \citep[e.g.][]{Ronquist2012mrbayes}.
However, when the Bayesian inference fails to recover a stable tree topology, the resulting consensus tree may contain many polytomies up to the maximum (resulting in a star tree, i.e. no resolution at all).
Because one cannot compare trees with no resolution we didn't use the consensus tree but instead we used the mode of 1000 Random Pairwise Bayesian Tree Comparisons (RPBTC).
The RPBTC allowed us to assess the difference between Bayesian trees posterior distribution by taking into account both the support value and the resolution of the consensus tree.
The RPBTC gives a low score if comparing any Bayesian trees posterior distribution to a Bayesian trees posterior distribution with a low resolution and a low node support (if any), even if the same posterior distribution is compared to itself.
Even though the RPBTC is dependent on 1000 random tree drawings, the output tree comparison score (i.e. the mode) is still significant not different between independent replicates (100 replicates of 1000 RPBTC each - see \hyperref[supplementaries]{supplementaries})
%
%
%
In other words, we summarized the difference between two Bayesian tree posterior distributions using the mode of the distribution of 1000 random trees drawn from the first Bayesian trees posterior distribution to 1000 random trees drawn from the the second Bayesian trees posterior distribution.
We used the mode to summarize the distributions because it reflects more accurately what a consensus tree represents (the topology the most represented in the posterior distribution).
A matrix with a good phylogenetic signal will lead to a posterior trees distribution containing many identical topologies.
Using a normalised tree similarity metric to compare the trees within such a posterior distribution will lead to a high value of tree similarity, reflecting reality.
Alternatively, a matrix with a poor phylogenetic signal will lead to a posterior distribution containing many different topologies.
Using a normalised tree similarity metric to compare the trees within such a posterior distribution will lead to a low value of tree similarity, reflecting reality. %Add something in the supplementaries or a figure... Don't know yet what.
We compared the pairs of Bayesian tree posterior distributions in the same way as for the Maximum Likelihood trees ("True" and "Best" tree compared to the tree "missing-data" tree).
This resulted in a distribution of one mode per chain for each "missing-data" trees (i.e. 50 modes) for which we calculated the mode and the 50\% and 95\% confidence intervals (see \hyperref[supplementaries]{supplementaries} - Fig. ~\ref{Fig_Compare}).

@@

\subsection{Random Bayesian tree comparisons}
We assessed the power of the Random Pairwise Bayesian Tree Comparison (RPBTC) method by comparing 1000 random trees from a posterior distribution trees set to another 1000 random trees from the same posterior distribution trees set.
We repeated this 100 times independently using the same posterior distribution trees set each time resulting in 100 replicates of the same posterior distribution trees set compared 1000 times.
We used an anova to test if there was no significant difference between the replicates so that the RBTC can be replicated.
We applied this protocol on a poorly resolved tree, a resolved tree with low support value and a resolved tree with high support values.
%Insert table with results here.

>>>>

For the Bayesian trees, we had to transform the trees posterior distribution to a single value in order to be able to compare it to the results of the Maximum Likelihood trees comparisons.
Traditionally people use the 50\% consensus rule in order to summarise a Bayesian tree posterior distribution \citep[e.g.][]{Ronquist2012mrbayes}.
However, when the Bayesian inference fails to recover a stable tree topology, the resulting consensus tree may contain many polytomies up to the maximum (resulting in a star tree, i.e. no resolution at all).
Because one cannot compare trees with no resolution we didn't use the consensus tree but instead we used the mode of 1000 Random Pairwise Bayesian Tree Comparisons (RPBTC).
The RPBTC allowed us to assess the difference between Bayesian trees posterior distribution by taking into account both the support value and the resolution of the consensus tree.
The RPBTC gives a low score if comparing any Bayesian trees posterior distribution to a Bayesian trees posterior distribution with a low resolution and a low node support (if any), even if the same posterior distribution is compared to itself.
Even though the RPBTC is dependent on 1000 random tree drawings, the output tree comparison score (i.e. the mode) is still significant not different between independent replicates (100 replicates of 1000 RPBTC each - see \hyperref[supplementaries]{supplementaries}, table ~\ref{RPBTC_testing})
%
%
%
In other words, we summarized the difference between two Bayesian tree posterior distributions using the mode of the distribution of 1000 random trees drawn from the first Bayesian trees posterior distribution to 1000 random trees drawn from the the second Bayesian trees posterior distribution.
We used the mode to summarize the distributions because it reflects more accurately what a consensus tree represents (the topology the most represented in the posterior distribution).
A matrix with a good phylogenetic signal will lead to a posterior trees distribution containing many identical topologies.
Using a normalised tree similarity metric to compare the trees within such a posterior distribution will lead to a high value of tree similarity, reflecting reality.
Alternatively, a matrix with a poor phylogenetic signal will lead to a posterior distribution containing many different topologies.
Using a normalised tree similarity metric to compare the trees within such a posterior distribution will lead to a low value of tree similarity, reflecting reality. %Add something in the supplementaries or a figure... Don't know yet what.
We compared the pairs of Bayesian tree posterior distributions in the same way as for the Maximum Likelihood trees ("True" and "Best" tree compared to the tree "missing-data" tree).
This resulted in a distribution of one mode per chain for each "missing-data" trees (i.e. 50 modes) for which we calculated the mode and the 50\% and 95\% confidence intervals (see \hyperref[supplementaries]{supplementaries} - Fig. ~\ref{Fig_Compare}).

@@

\subsection{Random Bayesian tree comparisons}
We assessed the power of the Random Pairwise Bayesian Tree Comparison (RPBTC) method by comparing 1000 random trees from a posterior distribution trees set to another 1000 random trees from the same posterior distribution trees set.
We repeated this 100 times independently using the same posterior distribution trees set each time resulting in 100 replicates of the same posterior distribution trees set compared 1000 times.
We used an anova to test if there was no significant difference between the replicates so that the RBTC can be replicated.
We applied this protocol on a poorly resolved tree (Low Score), a resolved tree with low support value (Medium Score) and a resolved tree with high support values (High Score).
Results are available in table ~\ref{RPBTC_testing}.

\begin{table}[ht]
\caption{Anova results: difference between 100 replicates using the RPBTC method}
\centering
\begin{tabular}{rllrrrr}
  \hline
  Tree.Type & Used.metric & Replicates & Df & F.value & p.value \\ [0.5ex]
  \hline
  Low Score & RF & 100.00 &   1 & 0.07 & 0.79 \\ [0.5ex]
  Low Score & Tr & 100.00 &   1 & 0.33 & 0.56 \\ [0.5ex]
  Medium Score & RF & 100.00 &   1 & 1.07 & 0.30 \\ [0.5ex]
  Medium Score & Tr & 100.00 &   1 & 0.00 & 0.96 \\ [0.5ex]
  High Score & RF & 100.00 &   1 & 0.20 & 0.66 \\ [0.5ex]
  High Score & Tr & 100.00 &   1 & 0.61 & 0.43 \\ [0.5ex]
  \hline
\end{tabular}
\label{RPBTC_testing}
\end{table}
